Simple Approximation Algorithms for Partitional and Hierarchical Clustering

### 2015-05-10 tex/project.tex
Last edits from me

\twocolumn[
\icmltitle{Local Search Approximation Algorithms for Partitional and Hierarchical Clustering}

@@

There are two main branches of clustering: partitional (or flat) and hierarchical. In partitional clustering we want to partition the data into disjoint sets whereas in hierarchical clustering we are interested in finding a nested tree of partitions. Often these problems are solved using heuristic algorithms which have no guarantee on the solution produced but are typically fast and easy to implement. We will look at how local search approximation algorithms, which do have some approximation guarantee on the constructed solution and are less computationally expensive than other approximation methods, compare in practice to some of the most widely used clustering algorithms both in the quality of the solution and in computation time.

@@

More recently, spectral based methods have become popular for clustering. These algorithms construct a graph based on the similarity matrix and compute the eigenvectors of the normalized graph Laplacian matrix \cite{Luxburg}. These eigenvectors are used to split the current cluster in half using the second eigenvector (continuing recursively until there are $k$ clusters).  Another method is to use the eigenvectors to project the data points into a lower dimensional space on which the $k$-means algorithm is run \cite{Verma}. We will base our comparisons against the algorithm introduced by \cite{Meila} which uses the latter method and performed best in practice on several datasets \cite{Verma}.

@@

Various approximation algorithms for the $k$-median problem have been proposed, with a best known approximation ratio of 2.592 \cite{Wu}. However, this algorithm uses linear programming and is computationally intensive compared to local search and greedy algorithms.  It is known that a local search algorithm that considers swaps of size $\leq p$ is a $(3 + \frac{2}{p})$-approximation algorithm \cite{Arya}. Note that for $p=1$, this is equivalent to PAM.  As far as we know, this algorithm has not been implemented or tested on clustering data, so the relative trade-off of solution quality and computation time as $p$ increases is unknown.

@@

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the flat partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. The local search $(3+\frac{2}{p})$-approximation algorithm as the $k$-median solver performed well in experimental tests \cite{Nagarajan}. While the authors tested the variants of this algorithm against other $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

@@

\subsection{Local Search with Multiswaps}
Consider a local search algorithm which starts with a random subset $S$ of $k$ points. At every iteration, the algorithm looks for a set $A \subseteq S$ and a set $B \subseteq X -S$ such that $|A| = |B| \leq p$ and
\[ f(S- B \cup A) \geq f(S). \]
If the algorithm finds such a set, it swaps out $B$ from $S$ and adds $A$. If no such set exists, the algorithm outputs $S$. At every iteration, for $p$ independent of $k$ and $n$, let $\mathcal{N}$ be the size of the search neighborhood. If we ensure that the algorithm improves the objective by at least a factor of $(1- \delta) = (1- \varepsilon/\mathcal{N})$, this is a polynomial-time algorithm. Furthermore, the local search algorithm outputs $S \subseteq X$ such that $|S| = k$ and
\[ f(S) \leq \frac{1}{1-\varepsilon} \left (3 + \frac{2}{p} \right ) f(O) \]
for all $O \subseteq X$. Therefore, there is a trade-off between the computation time of this algorithm and the guaranteed quality of the solution.

@@

Note that $\psi$ requires summing an exponential number of terms. However, we can estimate $h(A)$ (and therefore $\psi(A)$) by randomly sampling $g(B)$. If $\hat{\psi}(A)$ is an estimate of $\psi(A)$ sampled from $\Omega (\varepsilon^{-2} n^4 \ln^4 n \ln M)$ samples, then 
\[ Pr[ | \hat{\psi}(A) - \psi(A) | \geq \delta ] = O(M^{-1}). \]
By setting $M$ large enough (polynomially large), the algorithm will converge in polynomial time and will return a solution satisying the above approximation ratio with high probability \cite{Sviridenko}.

@@

It is interesting to note that higher dimensional Gaussian data, with smaller values of $k$ and a lower percentage of noisy data, was easier for the algorithms to classify, which makes sense based on our intuition of clustering. For the OR library instances, we saw that the approximation ratio tended to increase as $k$ grew larger but not as $n$ grew.

@@

From each non-empty bucket, pick the solution with the smallest number of cluster centers and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r = V_n$. We then inductively build our nested solutions. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we define $S_i$ by iteratively adding the closest point $j \in S_{i-1}$ to $i$ for all $i \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

@@

\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & F-Score & F-Score & F-Score & F-Score & F-Score & F-Score \\ \hline
	Dataset & Ave Linking & Ward Linking & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 0.615 & 0.510 & 0.772 & 0.643 & 0.742 & 0.633 \\ 
	Iris & 0.5 & 0.5 & 0.888 & 0.767 & 0.762 & 0.732 \\ 
	Gauss\_5\_5\_5 & 0.519 & 0.516 & 0.591 & 0.816 & 0.761 & 0.785 \\
	Gauss\_5\_10\_5 & 0.298 & 0.289 & 0.589 & 0.584 & 0.576 & 0.446 \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.393 & 0.375 & 0.649 & 0.616 & 0.624 & 0.624 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | }
\hline
	   & Approx & Approx & Approx & Approx \\ \hline
	Dataset & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 1.585 & 2.334 & 1.711 & 2.001 \\ 
	Iris & 2560631.459 & 2550279.566 & 2787312.009 & 1623132.096 \\ 
	Gauss\_5\_5\_5 & 6.459 & 13.394 & 8.624 & 7.791 \\
	Gauss\_5\_10\_5 &  3.420 & 3.756 & 4.196 & 2.309  \\
	%Gauss\_5\_50\_5 &  & & &  \\
	Gaussian Average & 5.182 & 7.262  & 5.537 & 4.739 \\ 
	pmed5 & 11.367 & 51.141 & 16.261 & 11.950 \\ 
	pmed25 & 6.641 & 13.372 & 5.378 & 13.200 \\ 
	pmedian avg & 10.330 & 19.822 & 10.524 & 8.599 \\ \hline
\end{tabular}

\caption{Average Computation Time (sec), FScore measure, and Approximation Coeffiecients for Hierarchical Clustering Algorithms}
\end{figure*}

@@

%\section{Acknowledgements}
%We would like to acknowledge David P. Williamson and Karthik Sridharan for their valuable feedback on this project.

>>>>

\twocolumn[
\icmltitle{Simple Approximation Algorithms for Partitional and Hierarchical Clustering}

@@

There are two main branches of clustering: partitional (or flat) and hierarchical. In partitional clustering we want to partition the data into disjoint sets whereas in hierarchical clustering we are interested in finding a nested tree of partitions. Often these problems are solved using heuristic algorithms which have no guarantee on the solution produced but are typically fast and easy to implement. We will look at how local search and greedy approximation algorithms, which do have some approximation guarantee on the constructed solution and are less computationally expensive than other approximation methods, compare in practice to some of the most widely used clustering algorithms both in the quality of the solution and in computation time.

@@

More recently, spectral based methods have become popular for clustering. These algorithms construct a graph based on the similarity matrix and compute the eigenvectors of the normalized graph Laplacian matrix \cite{Luxburg}. These eigenvectors are used to split the current cluster in half using the eigenvector associated with the second highest eigenvalue (continuing recursively until there are $k$ clusters).  Another method is to use the eigenvectors to project the data points into a lower dimensional space on which the $k$-means algorithm is run \cite{Verma}. We will base our comparisons against the algorithm introduced by \cite{Meila} which uses the latter method and performed best in practice on several datasets \cite{Verma}.

@@

The $k$-median problem is also a fundamental problem in approximation algorithms. However, these algorithms are often not tested in practice despite their improved guarantees on the solution. The best known approximation ratio for this problem is 2.592 \cite{Wu}. This algorithm uses linear programming and is computationally intensive compared to local search and greedy algorithms.  It is known that a local search algorithm that considers swaps of size $\leq p$ is a $(3 + \frac{2}{p})$-approximation algorithm \cite{Arya}. Note that for $p=1$, this is equivalent to PAM.  As far as we know, this algorithm has not been implemented or tested on clustering data, so the relative trade-off of solution quality and computation time as $p$ increases is unknown.

@@

Approximation algorithms for hierarchical clustering have also been studied, where we aim to identify a hierarchy of clusters so as to minimize the ratio of the objective value on every level in the hierarchy of size $k$ to the optimal objective value of the flat partitional clustering problem with $k$ clusters. Note that it may not always be possible to find a solution with approximation ratio 1. The hierarchical $k$-median problem was originally introduced by Plaxton who gave a 238.88 approximation algorithm \cite{Plaxton}. This was improved by Lin et al.~\cite{Lin} who gave an incremental approximation algorithm framework. In particular, their approach relies on a black box $k$-median solver and then uses these solutions for $k=1, \ldots, n$ to greedily construct a hierarchical clustering that increases the objective value by at most a factor of $20.71$ (or $10.03$ when randomized) at each level. Using the local $p$-swap $(3+\frac{2}{p})$-approximation algorithm mentioned above as the black-box $k$-median solver performed well in experimental tests \cite{Nagarajan}. While the authors tested the variants of this algorithm against other $k$-median approximation algorithms, it has not been tested against traditional clustering algorithms or on any labeled data sets.

@@

\subsection{Local Search with Multiswaps}
Consider a local search algorithm which starts with a random subset $S$ of $k$ points. At every iteration, the algorithm looks for a set $B \subseteq S$ and a set $A \subseteq X -S$ such that $|A| = |B| \leq p$ and
\[ f(S- B \cup A) \geq f(S). \]
If the algorithm finds such sets, it swaps out $B$ from $S$ and adds $A$. If no such set exists, the algorithm outputs $S$. At every iteration, for $p$ independent of $k$ and $n$, let $\mathcal{N}$ be the size of the search neighborhood. If we ensure that the algorithm improves the objective by at least a factor of $(1- \delta) = (1- \varepsilon/\mathcal{N})$, this is a polynomial-time algorithm. Furthermore, the local search algorithm outputs $S \subseteq X$ such that $|S| = k$ and
\[ f(S) \leq \frac{1}{1-\varepsilon} \left (3 + \frac{2}{p} \right ) f(O) \]
for all $O \subseteq X$. Therefore, there is a trade-off between the computation time of this algorithm and the guaranteed quality of the solution.

@@

Note that $\psi$ requires summing an exponential number of terms. However, we can estimate $h(A)$ (and therefore $\psi(A)$) by randomly sampling $g(B)$. If $\hat{\psi}(A)$ is an estimate of $\psi(A)$ sampled from $\Omega (\varepsilon^{-2} n^4 \ln^4 n \ln M)$ samples, then 
\[ Pr[ | \hat{\psi}(A) - \psi(A) | \geq \delta ] = O(M^{-1}). \]
By setting $M$ large enough (polynomially large), the algorithm will converge in polynomial time and will return a solution satisying the above approximation ratio with high probability \cite{Sviridenko}. As far as we are aware, this algorithm has not been tested for any clustering problems.

@@

It is interesting to note that, for the Gaussian data, an increase in dimension, reduction in $k$, or reduction in the percentage of noisy data made it easier for the algorithms to classify the data set, which makes sense based on our intuition of clustering. For the OR library instances, we saw that the approximation ratio tended to increase as $k$ grew larger but not as $n$ grew.

@@

From each non-empty bucket, pick the solution with the smallest number of cluster centers, and let these solutions be $\overline{V}_1, \overline{V}_2, \ldots \overline{V}_r$. We then inductively build our nested solutions. Given $(S_{i-1}, a_{i-1})$ and $\overline{V}_i$, we define $S_i$ by iteratively adding the closest point $j \in S_{i-1}$ to $k$ for all $k \in \overline{V}_i$. Note that $|S_{i}| \leq |\overline{V}_i|$.

@@

\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & F-Score & F-Score & F-Score & F-Score & F-Score & F-Score \\ \hline
	Dataset & Ave Linking & Ward Linking & K-med & K-med logn rand & K-med rand & PAM logn rand \\ \hline
	Soybean & 0.615 & 0.510 & 0.772 & 0.643 & 0.742 & 0.633 \\ 
	Iris & 0.5 & 0.5 & 0.888 & 0.767 & 0.762 & 0.732 \\ 
	Gauss\_5\_5\_5 & 0.519 & 0.516 & 0.591 & 0.816 & 0.761 & 0.785 \\
	Gauss\_5\_10\_5 & 0.298 & 0.289 & 0.589 & 0.584 & 0.576 & 0.446 \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.393 & 0.375 & 0.649 & 0.616 & 0.624 & 0.624 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | }
\hline
	   & Approx & Approx & Approx & Approx \\ \hline
	Dataset & K-med & K-med logn rand & K-med rand & PAM logn rand \\ \hline
	Soybean & 1.585 & 2.334 & 1.711 & 2.001 \\ 
	Iris & 2560631.459 & 2550279.566 & 2787312.009 & 1623132.096 \\ 
	Gauss\_5\_5\_5 & 6.459 & 13.394 & 8.624 & 7.791 \\
	Gauss\_5\_10\_5 &  3.420 & 3.756 & 4.196 & 2.309  \\
	%Gauss\_5\_50\_5 &  & & &  \\
	Gaussian Average & 5.182 & 7.262  & 5.537 & 4.739 \\ 
	pmed5 & 11.367 & 51.141 & 16.261 & 11.950 \\ 
	pmed25 & 6.641 & 13.372 & 5.378 & 13.200 \\ 
	pmedian avg & 10.330 & 19.822 & 10.524 & 8.599 \\ \hline
\end{tabular}

\caption{Average Computation Time (sec), FScore measure, and Approximation Coeffiecients for Hierarchical Clustering Algorithms. K-med and PAM represent the black-box $k$-median solvers used and rand and logn indicate whether the randomized and logarithmic setting were used, respectively.}
\end{figure*}

@@

\section{Acknowledgements}
We would like to acknowledge David P. Williamson and Karthik Sridharan for their valuable feedback on this project.
